Tulai

### 2023-09-12 paper/paper_tulai.qmd
edited most parts-made main figures for our purpose

             TRUE ~ as.character(depth)
  )) %>% 
  separate(depth, 
           into = c("upper",
                    "lower"),
           sep = "-",
           remove = FALSE,
           convert = TRUE) %>% 
  mutate(lower = parse_number(lower)) %>% 
  rowwise() %>% 
  mutate(midpoint = mean(c(upper, lower))) %>%
# create 'layer' column
  mutate(layer = case_when(
      area == "tp1" & depth == "0-10 cm" ~ "c",
      area == "tp1" & depth == "0-20 cm" ~ "c",
      area == "tp1" & depth == "10-30 cm" ~ "c",
      area == "tp1" & depth == "10-90 cm" ~ "c",
      area == "tp1" & depth == "20-30 cm" ~ "c",
      area == "tp1" & depth == "30-40 cm" ~ "c",
      area == "tp1" & depth == "40-50 cm" ~ "c",
      area == "tp1" & depth == "50-100 cm" ~ "c",
      area == "tp1" & depth == "50-60 cm" ~ "c",
      area == "tp1" & depth == "50-70 cm" ~ "c",
      area == "tp1" & depth == "80-90 cm" ~ "c",
      area == "tp1" & depth == "90-100 cm" ~ "c",
      area == "tp1" & depth == "100-110 cm" ~ "c",
      area == "tp1" & depth == "110-120 cm" ~ "c",
      area == "tp1" & depth == "120-130 cm" ~ "c",
      area == "tp1" & depth == "120-140 cm" ~ "c",
      area == "d1" & depth == "0-20 cm" ~ "c",
      area == "d1" & depth == "20-30 cm" ~ "c",
      area == "d1" & depth == "40-60 cm" ~ "a",
      area == "d1" & depth == "50-60 cm" ~ "a",
      area == "d1" & depth == "60-70 cm" ~ "a",
      area == "d1" & depth == "70-80 cm" ~ "a",
      area == "ta" & depth == "0-30 cm" ~ "e",
      area == "ta" & depth == "30-40 cm" ~ "e",
      area == "ta" & depth == "40-50 cm" ~ "e",
      area == "ta" & depth == "50-60 cm" ~ "e",
      area == "ta" & depth == "60-70 cm" ~ "e",
      area == "ta" & depth == "70-120 cm" ~ "d",
      area == "ta" & depth == "120-150 cm" ~ "d",
      area == "ta" & depth == "150-180 cm" ~ "d",
      area == "ta" & depth == "180-190 cm" ~ "d",
      area == "ta" & depth == "190-200 cm" ~ "d",
      area == "ta" & depth == "200-210 cm" ~ "d",
      area == "ta" & depth == "210-220 cm" ~ "d",
      area == "ta" & depth == "220-230 cm" ~ "d",
      area == "ta" & depth == "230-240 cm" ~ "d",
      area == "ta" & depth == "230-250 cm" ~ "d",
      area == "ta" & depth == "250-260 cm" ~ "d",
      area == "ta" & depth == "260-270 cm" ~ "d",

@@

```{r}
# BM: this is an unusual coding pattern, I'm curious about where you got it from?
tl3 <- tl3 %>%
  mutate(
    Blank2 = case_when(
      Blank2 == "flake" ~ "flake",  # Keep "flake" as it is
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` < 12 ~ "bladelet",
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` >= 12 ~ "blade",
      TRUE ~ Blank2
    )
  )  %>%
  mutate(
    Blank2 = case_when(
      Blank2 %in% c("blade", "bladelet", "flake") ~ Blank2,
      TRUE ~ NA_character_
    )
  )

@@

# BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:

@@

tl4 <- tl4 %>%
  # BM: the problem with this combination of mutate_all and ifelse that you use below is that they 
  # coerce all columns to be the same data type, so you'll see that all the numeric
  # columns are now character columns after you run these lines. So you need to
  # coerce those number columns back to numeric type so you can plot and compute
  # with them
  mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
  # BM: what does this next line do?
  mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
  # BM: what is this SHINE variable? I'm not familiar with it
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

@@

```{r}
# convert all columns to character type (I DONT KNOW WHY I DID THIS!)
# BM: my preference is not to see comments and code like this. Go ahead and delete it.
tl4 <- tl4 %>%
  mutate(across(everything(), as.character))

# replace 'NA', 'na', or '0' with R's NA
tl4 <- tl4 %>%
  # BM: in my opinion this is nicer than mutate_all and ifelse
  mutate(across(everything(), ~na_if(., "NA"))) %>%
  mutate(across(everything(), ~na_if(., "na"))) %>%
  mutate(across(everything(), ~na_if(., "0")))

@@

```{r}
#set the value of utilization to NA only for the rows where `sickle shine` == 1. 
#All other rows should remain unchanged

#unsuccessful, should try again later

@@

@@

# BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
knitr::kable(summary_table_area)

@@

```{r}
# Create a summary table
summary_table_layer <- tl_final %>%
  group_by(layer) %>%  # Note that 'area' is replaced with 'layer'
  summarise(
    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

@@

DEPTH:

```{r}
# Create a summary table based on the 'depth' column
summary_table_depth <- tl_final %>%
  mutate(depth = gsub("cm", "", depth)) %>%
  group_by(depth) %>%  
  summarise(
    `Cores (n)` = sum(!is.na(core)),  
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

kable(summary_table_depth, 
      caption = "Table: Summary by Depth", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

```{r}
# Create a summary table based on the 'depth', 'area', and 'layer' columns
summary_table_depth_area_layer <- tl_final %>%
  group_by(depth, area, layer, upper, lower) %>%  
  summarise(
    `Cores (n)` = sum(!is.na(core)),  
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

@@

```{r}
# Prepare data for Summary table-al
summary_table_al <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>% 
  group_by(al) %>% 
  summarise(
    blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
    total = n()
  )

# Adding a new row with total counts
summary_table_al <- summary_table_al %>% 
  add_row(
    al = "Total",
    blade_count =              sum(.$blade_count, na.rm = TRUE),
    flake_count =              sum(.$flake_count, na.rm = TRUE),
    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
    tool_count =               sum(.$tool_count, na.rm = TRUE),
    core_count =               sum(.$core_count, na.rm = TRUE),
    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
    total =                    sum(.$total, na.rm = TRUE)  # Explicitly setting the total
  )

```{r}
summary_table_al <- summary_table_al %>%
  filter(!(al %in% c("D1_NA", "TP1_NA")))

@@

```{r}
#summary table-area

# Rename columns
summary_table_area <- summary_table_area %>%
  rename(
    "Area" = area,
    "Blade" = blade_count,
    "Flake" = flake_count,
    "Bladelet" = bladelet_count,
    "Retouched Tools" = retouched_tool_count,
    "Tools" = tool_count,
    "Utilized Tools" = utilized_tool_count,
    "Sickle Shine" = shine_tool_count,
    "Cores" = core_count,
    "Core rejuvenations pieces" = core_rejuvenation_count,
    "Total" = total
  ) %>%
  mutate(
    Area = if_else(row_number() == n() & is.na(Area), "Total", Area)
  ) %>%
  mutate(
    Area = recode(Area, "ta" = "TA", "d1" = "D1", "tp1" = "TP1", "Total" = "Total")
  )

kable(summary_table_area,
      caption = "Table 1: Count of debitage type and tools in three different main areas.",
      align = 'c') %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    font_size = 12
  ) %>%
  row_spec(0, font_size = 11)
```

```{r}
# summary table-core and core rejuvenation
```

```{r}
# summary table for tools: retouched-utelized-sickle

# Create a summary table
summary_table_tool_area <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  # Exclude rows where 'tooltypegroup' is NA or empty
  group_by(area, tooltypegroup) %>%  # Group the data by area and tooltypegroup
  tally() %>%  # Count the number of rows in each group
  arrange(area, tooltypegroup) %>%  # Arrange by area and tooltypegroup
  spread(key = area, value = n, fill = 0)  # Spread the 'area' column into multiple columns

@@

# BM: looks like you need this data starting from around here in the document
library(googlesheets4)

@@

sum_depth_sheet <- sum_depth_sheet %>%
  mutate(`depth (cm)` = unlist(`depth (cm)`),
         level = unlist(level)) %>% 
  mutate(area_layer_level = paste0(toupper(area), 
                                   "_", 
                                   layer, 
                                   "_", 
                                   level)) %>%
  relocate(area_layer_level, 
           .after = last_col()) %>%
  mutate(
    retouch_freq = round(retouch_tool / total_lithics, 2),
    lithic_dens = round(total_lithics / volume, 2)
  )

@@

volume for each area:

```{r}
# Define the volumes for each area 
volumes <- data.frame(
  area = c("ta", "tp1", "d1"),
  volume = c(52, 5.76, 2.1375) # in m^3
  # BM: My preference would be to show the calculation of these 
  # values from the raw data, rather than type them in 
  # here, where we can't see where they came from, we can't easily
  # verify these values as they are currently 
)

>>>>

             TRUE ~ as.character(depth)
  )) %>%

  separate(depth, 
           into = c("upper",
                    "lower"),
           sep = "-",
           remove = FALSE,
           convert = TRUE) %>% 
  mutate(lower = parse_number(lower)) %>% 
  rowwise() %>% 
  mutate(midpoint = mean(c(upper, lower))) %>%
# create 'layer' column
  mutate(layer = case_when(
      area == "tp1" & depth == "0-10 cm" ~ "c",
      area == "tp1" & depth == "0-20 cm" ~ "c",
      area == "tp1" & depth == "10-30 cm" ~ "c",
      area == "tp1" & depth == "10-90 cm" ~ "c",
      area == "tp1" & depth == "20-30 cm" ~ "c",
      area == "tp1" & depth == "30-40 cm" ~ "c",
      area == "tp1" & depth == "40-50 cm" ~ "c",
      area == "tp1" & depth == "50-100 cm" ~ "c",
      area == "tp1" & depth == "50-60 cm" ~ "c",
      area == "tp1" & depth == "50-70 cm" ~ "c",
      area == "tp1" & depth == "80-90 cm" ~ "c",
      area == "tp1" & depth == "90-100 cm" ~ "c",
      area == "tp1" & depth == "100-110 cm" ~ "c",
      area == "tp1" & depth == "110-120 cm" ~ "c",
      area == "tp1" & depth == "120-130 cm" ~ "c",
      area == "tp1" & depth == "120-140 cm" ~ "c",
      area == "d1" & depth == "0-20 cm" ~ "c",
      area == "d1" & depth == "20-30 cm" ~ "c",
      area == "d1" & depth == "40-60 cm" ~ "a",
      area == "d1" & depth == "50-60 cm" ~ "a",
      area == "d1" & depth == "60-70 cm" ~ "a",
      area == "d1" & depth == "70-80 cm" ~ "a",
      area == "ta" & depth == "0-30 cm" ~ "e",
      area == "ta" & depth == "30-40 cm" ~ "e",
      area == "ta" & depth == "40-50 cm" ~ "e",
      area == "ta" & depth == "50-60 cm" ~ "e",
      area == "ta" & depth == "60-70 cm" ~ "e",
      area == "ta" & depth == "70-120 cm" ~ "d",
      area == "ta" & depth == "120-150 cm" ~ "d",
      area == "ta" & depth == "150-180 cm" ~ "d",
      area == "ta" & depth == "180-190 cm" ~ "d",
      area == "ta" & depth == "190-200 cm" ~ "d",
      area == "ta" & depth == "200-210 cm" ~ "d",
      area == "ta" & depth == "210-220 cm" ~ "d",
      area == "ta" & depth == "220-230 cm" ~ "d",
      area == "ta" & depth == "230-240 cm" ~ "d",
      area == "ta" & depth == "230-250 cm" ~ "d",
      area == "ta" & depth == "250-260 cm" ~ "d",
      area == "ta" & depth == "260-270 cm" ~ "d",

@@

```{r}
# BM: this is an unusual coding pattern, I'm curious about where you got it from?

tl3 <- tl3 %>%
  mutate(
    Blank2 = case_when(
      Blank2 == "flake" ~ "flake",  # Keep "flake" as it is
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` < 12 ~ "bladelet",
      (`Core Typology` %in% c("0", "NA", "na", "-", ".") | 
       is.na(`Core Typology`) | 
       `Core Typology` == "") & `Width (mm)` >= 12 ~ "blade",
      TRUE ~ Blank2
    )
  )  %>%
  mutate(
    Blank2 = case_when(
      Blank2 %in% c("blade", "bladelet", "flake") ~ Blank2,
      TRUE ~ NA_character_
    )
  )

@@

# BM: why are we removing these rows? A comment here would be good to explain. A simpler way, by referencing the row numbers directly:
# SS: they are natural rocks, and don't want to count them as lithic

@@

tl4 <- tl4 %>%
  # BM: the problem with this combination of mutate_all and ifelse that you use below is that they 
  # coerce all columns to be the same data type, so you'll see that all the numeric
  # columns are now character columns after you run these lines. So you need to
  # coerce those number columns back to numeric type so you can plot and compute
  # with them
  mutate_all(~ ifelse(is.na(.x) | .x == "", NA, .x)) %>% 
  # BM: what does this next line do?
  #SS: I wanted to remove any asterisks, whitespace, or *, ?, or !. is that wrong?

  mutate_all(~ str_replace(., "(?<=\\D)\\*(?=\\D)|\\s+|[*?!]+", "")) %>%
  # BM: what is this SHINE variable? I'm not familiar with it

  ##SS: we meant sickle shine, this is one of the criteria based on which Hole claimed the site is not related to agriculture (similar to ALi Kosh located in Dehluran)
  mutate(
    SHINE = str_trim(SHINE) %>% 
            str_to_title(),
    shine2 = SHINE
  ) %>%
  mutate(
    shine2 = case_when(
      SHINE %in% c("Subparallel", 
                   "Parallel", 
                   "Sub-Parallel", 
                   "Parallel-Subprarallel",
                   "Scaled", 
                   "Sub-Paralel", 
                   "Sub-Parallell", 
                   "Sub-Paallel",
                   "Sub-Paralell", 
                   "Sub-Prallel", 
                   "L:semiparallel/R:scaled",
                   "Semiparallel", 
                   "Semi-Parallel", 
                   "Scaled/Sub-Parallel",
                   "Seb-Parallel", 
                   "Semi-Abrupt") ~ NA_character_,
      TRUE ~ SHINE
    ),
    Utilization = if_else(Utilization == "0", NA_character_, Utilization),
    Retouch =     if_else(Retouch == "0", NA_character_, Retouch),
    SHINE =       if_else(SHINE == "0", NA_character_, SHINE),
    shine2 =      if_else(shine2 == "0", NA_character_, shine2),
    Blank2 =      if_else(Blank2 == "0", NA_character_, Blank2)
  ) %>% 
  mutate(
    SHINE = na_if(SHINE, "")
  )

@@

@@

```{r}
# List of columns that should be numeric
columns_to_convert <- c("cortex", 
                        "upper", 
                        "lower", 
                        "midpoint", 
                        "length", 
                        "width", 
                        "utilization", 
                        "retouch",
                        "sickle shine")

# Convert each column to numeric
for (col_name in columns_to_convert) {
  # Replace non-numeric characters with NA
  tl_final[[col_name]][tl_final[[col_name]] %in% c("", "NA", "N/A")] <- NA
  # Convert to numeric
  tl_final[[col_name]] <- as.numeric(tl_final[[col_name]])
}

# check the structure to see if they have been converted
str(tl_final)
```

```{r}
#set the value of utilization to NA only for the rows where `sickle shine` == 1.

tl_final <- tl_final %>%
  rowwise() %>%
  mutate(utilization = ifelse(is.na(`sickle shine`) | is.na(utilization), utilization,
                              ifelse(`sickle shine` == 1 & utilization == 1, NA_real_, utilization))) %>%
  ungroup()

```{r}
#number of `sickle shine` in clumn `sickle shine`=1 and tooltypegroup are not match: 
rev(sort(table(tl_final$tooltypegroup)))
rev(sort(table(tl_final$`sickle shine`)))

#troubleshoot:
tl_final %>% filter(`sickle shine` == 1) %>% 
  select(`tool-typology`, 
         `sickle shine`)

tl_final %>% filter(`tool-typology` == "sickle shine") %>% 
  select(`tool-typology`, `sickle shine`)

tl_final <- tl_final %>%
  mutate(
    `tool-typology` = case_when(
      `sickle shine` == 1 ~ "sickle shine",
      utilization == 1 ~ "utilized tool",
      TRUE ~ `tool-typology`               
    )
  )

rev(sort(table(tl_final$`tool-typology`)))
```

@@

```{r}
library(dplyr)

# Filter the data to include only the areas 'ta', 'd1', and 'tp1'
tl_final_area <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1"))

@@

# BM: looks like this is area and layer, is that right? I see c1, d1, d2, d3, d4 etc.
# SS: those are different excavated areas, we mainly consider ta, tp1, and d1.
knitr::kable(summary_table_area)

@@

```{r}
# Create a summary table
summary_table_layer <- tl_final %>%
  group_by(layer) %>% 
  summarise(
    `Cores (n)` = sum(!is.na(core)),  # Count non-NA core entries
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  # Count non-NA corefrag entries
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

@@

@@

```{r}
# Prepare data for Summary table-area_layer(al)
summary_table_al <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>% 
  group_by(al) %>% 
  summarise(
    blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    tool_count = sum(retouch == 1, na.rm = TRUE) + sum(utilization == 1, na.rm = TRUE),
    core_count = sum(!is.na(`core-typology`), na.rm = TRUE),
    core_rejuvenation_count = sum(!is.na(`corefrag`), na.rm = TRUE),
    retouched_tool_count = sum(retouch == 1, na.rm = TRUE),
    utilized_tool_count = sum(utilization == 1, na.rm = TRUE),
    shine_tool_count = sum(`sickle shine` == 1, na.rm = TRUE),
    total = n()
  )

# Adding a new row with total counts
summary_table_al <- summary_table_al %>% 
  add_row(
    al = "Total",
    blade_count =              sum(.$blade_count, na.rm = TRUE),
    flake_count =              sum(.$flake_count, na.rm = TRUE),
    bladelet_count =           sum(.$bladelet_count, na.rm = TRUE),
    tool_count =               sum(.$tool_count, na.rm = TRUE),
    core_count =               sum(.$core_count, na.rm = TRUE),
    core_rejuvenation_count =  sum(.$core_rejuvenation_count, na.rm = TRUE),
    retouched_tool_count =     sum(.$retouched_tool_count, na.rm = TRUE),
    utilized_tool_count =      sum(.$utilized_tool_count, na.rm = TRUE),
    shine_tool_count =         sum(.$shine_tool_count, na.rm = TRUE),
    total =                    sum(.$total, na.rm = TRUE)  
  )

rev(sort(table(tl_final$tooltypegroup)))
```

```{r}
#explore tool type group

summary_table_tooltype <- tl_final %>%
  filter(tooltypegroup != 'utilized tool') %>%  # Exclude 'utilized tool'
  group_by(tooltypegroup) %>% 
  summarise(

```{r}
#generate a table

summary_table_al <- summary_table_al %>%
  filter(!(al %in% c("D1_NA", "TP1_NA")))

@@

```{r}
# Create a summary table for typology based on area

summary_table_tool_area <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(!is.na(tooltypegroup), tooltypegroup != "") %>%  # Exclude rows where 'tooltypegroup' is NA or empty
  group_by(area, tooltypegroup) %>%  # Group the data by area and tooltypegroup
  tally() %>%  # Count the number of rows in each group
  arrange(area, tooltypegroup) %>%  # Arrange by area and tooltypegroup
  spread(key = area, value = n, fill = 0)  # Spread the 'area' column into multiple columns

@@

```{r}
# Create a summary table based on the 'depth', 'area', and 'layer' columns
tl_final <- tl_final %>%
  mutate(depth = gsub("cm", "", depth))

summary_table_depth_area_layer <- tl_final %>%
  group_by(TA, depth, area, layer, upper, lower) %>%  
  summarise(
    `Cores (n)` = sum(!is.na(core)),  
    `Core rejuvenations pieces (n)` = sum(!is.na(corefrag)),  
    `Blade (n)` = sum(blank == "blade", na.rm = TRUE),
    `Flake (n)` = sum(blank == "flake", na.rm = TRUE),
    `Bladelet (n)` = sum(blank == "bladelet", na.rm = TRUE),
    `Retouched Tools (n)` = sum(retouch == 1, na.rm = TRUE),
    `Utilized (n)` = sum(utilization == 1, na.rm = TRUE),
    `Sickle Shine (n)` = sum(`sickle shine` == 1, na.rm = TRUE)
  )

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)
```

```{r}
#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.

# BM: looks like you need this data starting from around here in the document

library(tidyverse)
library(googlesheets4)

@@

@@

#SS: is that true?: (I have a weird number in row 5, cl Depth and couldn't find out what happened)

# Separate `excavation area` into `length` and `width` and calculate `volume`
sum_depth_sheet <- sum_depth_sheet %>%
  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
  mutate(
    length = as.numeric(length),
    width = as.numeric(width),
    `thickness of deposit` = `thickness of deposit` / 100  # Convert from cm to m
  ) %>%
  mutate(
    volume = length * width * `thickness of deposit`  # Calculate volume in m^3
  )

# Calculate lithic volumetric density 
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    `lithic_dens` = round(total_lithics / volume, 2)
  )

# Calculate retouch frequency 
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    retouch_freq = round(retouch_tool / volume, 2)  
  )

```{r}
#WABI plot: here I do not consider those levels with a value 0 of retouch frequency.

library(ggpmisc)

# Filter the data (get out those level with Retouch frequency=0)
filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)

# Create the plot: WABI_Level_logscale
WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels 
  geom_text(aes(label = paste(area, layer, level, sep = "_")), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

# Print the plot to display it
print(WABI_Level)

```{r}
#PCA_Level: preparing data

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should we not consider it?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
##core frequency
##blank frequency
##retouch frequency
##tool diversity

#Calculate Core Frequency
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    core_freq = round(cores / volume, 2)  
  )

# calculate Blank Frequency
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    blank_freq = round((flake + blade + bladelet) / volume, 2)  
  )

# calculate Tool Diversity
##calculating tool diversity is a bit tricky here; as I have some spitted depth, which we assign the number of artefacts based on the calculation of splitting, so how should I consider the tool diversity here?
```

PCA_al:

```{r}
#PCA: area_layer (al)
#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level. we have 5 stratigraphical layers: e, d, c, b, a
#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
#e: excavated area is A/ depth of excavation is 0-75(mean)
#d: excavated area is A/ 0.75m-3m→ 2.25
#c: d1 and TP → 30 cm and 40cm
#b: TP/ 170 cm 
#a: D1: 70cm

# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
library(dplyr)

summary_table_al_pca <- summary_table_al %>%
  mutate(
    length = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 4,
      al == "TP1_c" ~ 2,
      al == "TA_e" ~ 4,
      TRUE ~ NA_real_
    ),
    width = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 5,
      al == "TP1_c" ~ 1.8,
       al == "TA_e" ~ 5,
      TRUE ~ NA_real_
    ),
    thickness = case_when(
      al == "D1_a" ~ 0.7,
      al == "D1_c" ~ 0.3,
      al == "TA_d" ~ 2.25,
      al == "TP1_c" ~ 0.4,
      al == "TA_e" ~ 0.75,
      TRUE ~ NA_real_)
  ) %>%

  mutate(volume = length * width * thickness) %>%
  relocate(length, width, thickness, volume, .after = al)
# Show the updated data frame
summary_table_al_pca

```{r}
# Load required library
library(dplyr)

# Add a new column called 'blank_count'
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())

# Calculate core frequency, blank frequency, retouch frequency, lithic density 
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(
    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
    core_freq = round(core_count / volume, 2),
    blank_freq = round(blank_count / volume, 2),
    retouch_freq = round(retouched_tool_count / volume, 2)
  )

# Calculate tool diversity using Menhinick's index
summary_table_al_with_diversity <- tl_final %>% 
  filter(retouch == 1) %>%
  group_by(al, `tool-typology`) %>%  
  summarise(n = n(), .groups = 'drop') %>% 
  group_by(al) %>%  
  summarise(
    num_unique_tool_types = n_distinct(`tool-typology`),
    total_retouched_tools = sum(n),
    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
    .groups = 'drop'
  )

# Merge this new information into the existing summary_table_al_pca
summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")

# Remove the 'Total' row and select only the columns needed for PCA
fin_summary_table_al_pca <- summary_table_al_pca %>%
  filter(al != "Total") %>%
  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)

# Show the final table
print(fin_summary_table_al_pca)

```{r}
library(FactoMineR)
# Perform PCA
res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column

```{r}
## Got error: couldnt figure out the reason of error

# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
eigen_df <- data.frame(
  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
  Eigenvalue = round(eigenvalues[, 1], 3),
  `Variance percent` = round(eigenvalues[, 2], 3),
  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
)

# Print the table
print(eigen_df)

```{r}
# variables to PC1
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 1,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 2,  
             top = 10)

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 3,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 4,  
             top = 10)
```

```{r}
#Plot: Contribution of variables for each of the four relevant PCA dimensions
library(factoextra)
library(ggplot2)
